Meta-analytic approaches and effect sizes to account for 'nuisance heterogeneity' in comparative physiology

### 2021-10-08 MS/ms.Rmd
Full edit

In equation \@ref(eq:lnRR), $M_{1}$ is the mean of group 1 (e.g., a control group), whereas $M_{2}$ is the mean of group 2 (e.g., a treatment group). The mean, $M_{i}$, for group *i* can be any measurement type (e.g. a physiological rate, mass etc) so long as the variable is measured on the ratio scale. Natural log transformation of this ratio makes the effect size normally distributed (as commonly assumed by meta-analytic models). Equation \@ref(eq:slnRR) is the analytical solution for *lnRR*'s sampling variance where $SD_{1}^2$ and $SD_{2}^2$ are the sample variance in group 1 and 2, respectively, and $N_{1}$ and $N_{2}$ are the sample size in group 1 and 2.

@@

## **Meta-Analytic Models to Control for and Investigate Heterogeneous Effects Within and Across Studies**

It will not always be possible to derive an effect size that completely controls for nuisance heterogeneity at the effect size level. This limitation is partly because treatment effects may not be simple linear or exponential functions of the mean response, making the effect sizes we discuss above unsuitable [e.g., thermal performance curves; @Noble2018]. For example, dosages applied to treatments can follow quite complex non-linear relationships in relation to some mean response (e.g., quadratic relationships). When this occurs, we recommend comparative physiologists to apply meta-regression approaches, controlling for nuisance heterogeneity as a moderator. In many cases, this will be the easiest, or even preferred, approach because exploring drivers of heterogeneity is of prime interest in most meta-analyses anyway. Nonetheless, there are some simple statistical approaches a comparative physiologist can use to make the overall effects more easily interpretable, allowing one to control for nuisance heterogeneity, while also testing other biological moderators of interest [@Schielzeth2010]. Before diving into meta-regression models, we will first overview the meta-analytic models commonly used.

@@

A common goal of meta-analysis is to obtain an overall meta-analytic mean effect size estimate and some form of uncertainty around that mean, like a 95% confidence interval (CI). Including a prediction interval (PI) can also be useful (discussed below) [@Nakagawa2021c]. Estimating an overall mean and 95% CI is generally done with a simple random-effects or multilevel meta-analytic model that accounts for effect size sampling error [@Nakagawa2017; @NakagawaSantos2012]. More often than not, meta-analyses in comparative physiology will have complex hierarchical structure. For example, multiple effect sizes may be taken from a single study on traits measured on the same individuals or from many different species that share an evolutionary history [@Cinar2021; @Noble2017]. As such, it is far more likely that comparative physiologists will need to apply a multilevel meta-analytic model to control for the various sources of non-independence [@Cinar2021; @NakagawaSantos2012; @Song2021; @Nakagawa2021; @Noble2017]. Multilevel models are also extremely useful in describing the various sources contributing to effect size variance (a feature we describe below). For the purpose of our discussion, we will assume one needs to use more advanced multilevel models. Having said that, in some instances, it may be sufficient for a meta-analysis to make use of traditional random effects models. We refer readers to excellent reviews and books that overview the distinctions between common (fixed) effect, random-effect and multilevel models and when one might want (or not want) to apply these models [@Nakagawa2017; @NakagawaSantos2012; @Koricheva2013; @Borenstein2009; @Schmid2021].

@@

This model is a 'weighted' multilevel meta-analysis model because the estimation of the overall meta-analytic mean, $\beta_{0}$, is partially weighted by the inverse sampling variance of each effect size. Weighted meta-analytic models are important to use because they 1) improve precision on meta-analytic mean estimates by accounting for sampling variance; 2) allow for a formal analysis of heterogeneity; 3) and give precedence to higher quality studies (see below). From our survey, `r with(weighted, weighted[meta_analysis_weighted == "weighted model","percent"])*100`% of studies used a weighted meta-analytic model; however, of the studies self described as 'meta-analyses', `r with(weighted_meta, weighted_meta[meta_analysis_weighted == "unweighted model","percent"])*100`% (N = `r with(weighted_meta, weighted_meta[meta_analysis_weighted == "unweighted model","n"])` of `r with(weighted_meta, sum(weighted_meta[,"n"]))`) did not account for sampling variance. Many studies used 'Linear Regression' (including linear and generalised linear mixed effects models) as the main type of model fit to the data (Figure \@ref(fig:figsurv)C). Of the studies using weighted meta-analytic models `r with(het_reported, sum(het_reported[1:2,"percent"])*100)`% (N = `r with(het_reported, sum(het_reported[1:2,"n"]))` meta-analyses) did not report any measure of heterogeneity, which is unfortunate given how important these measures are to interpreting effects.

@@

#### *Multilevel Meta-regression to Understand Variation and Improve Interpretation*

Heterogeneity among effects that result from 'nuisance heterogeneity' can be dealt with using multilevel meta-regression models. Multilevel meta-regression models include all the same random effects that we have already discussed, but they also include fixed effects (i.e., predictors or moderators) that attempt to explain changes in the mean effect size. In other words, we can modify our multilevel meta-analytic model and turn it into a multilevel meta-regression model as follows:

@@

Note that this formula does not include $\sigma^2_{m}$ as sampling error variance is assumed to be known in meta-analysis, as explained above. When including continuous moderators, such as dosage, temperature, salinity, and even pH, $R^2_{marginal}$ provides a formal means to assess just how much variation in effects is the result of 'nuisance' heterogeneity at the effect size level. Other biological moderators of interest could also be included as fixed effects and $R^2_{marginal}$ can be calculated for each independently, or all together. Importantly, moderators could be those that explain variation at the effect size level (i.e., dosage, temperature) or variation at the study or species level (e.g., reproductive mode, endothermy, etc.). $R^2_{marginal}$, or how we like to think of it in this context as $R^2_{nuisance}$, maybe a useful measure to help readers understand just how much effect variation can result from the specific treatment application. If the heterogeneity is high as a result of, say, dosage differences, then it is clear that the choice of dosage will have a critical impact on the effect of interest. Comparative physiologists interested in implementing these calculations can do so using our packages `orchaRd` [@Nakagawa2021c] or `metaAidR` (<https://github.com/daniel1noble/metaAidR>).

@@

Comparative physiologists are interested in meta-analysing effects across a diversity of species, experimental designs, and environments. Importantly, sampling variances for many effect sizes commonly used by comparative physiologists (e.g., CTmax, $Q_{10}$ etc) can be formalized and powerful meta-analytic models can be easily applied to these effect measures (Figure \@ref(fig:fig0)). While an overall meta-analytic mean may be of interest, more often than not, understanding the drivers of effect size heterogeneity will be the primary interest. However, factors such as temperature and dosage differences across studies may generate obvious sources of heterogeneity that may not be of prime interest. Instead, comparative physiologists might simply want to focus on biological drivers of effect size variability and want an effect size or analytic approach that allows them to remove these ‘nuisances’. Surveying the comparative physiology literature shows that nuisance heterogenity is common, and is often not completely dealt with in a meta-analysis. Here, we provide a set of tools to deal with ‘nuisance heterogeneity’ (Figure \@ref(fig:fig1)) by reinventing standard effect sizes and/or using mean centered nuisance variables in meta-regression models. Estimating complex meta-regression models with a multitude of fixed effects might result, however, in more challenging model interpretation. As such, a combined approach, where some of the known heterogeneity is dealt with by using an appropriate effect size and associated sampling variance while other biologically interesting moderators are fit in a meta-regression model might be desired. Indeed, this may even be a necessity to simplify the model. The set of tools we describe provide clarification around importance and limitations of heterogeneity. We also hope to provide new ways to help comparative physiologists communicate effect measures more richly to guide our understanding and decisions around pressing global challenges.

>>>>

In equation \@ref(eq:lnRR), $M_{1}$ is the mean of group 1 (e.g., a control group), whereas $M_{2}$ is the mean of group 2 (e.g., a treatment group). The mean, $M_{i}$, for group *i* can be any measurement type (e.g. a physiological rate, mass etc) so long as the variable is measured on the ratio scale. Natural log transformation of this ratio makes the effect size normally distributed (as commonly assumed by meta-analytic models). Equation \@ref(eq:slnRR) is the analytical solution for *lnRR*'s sampling variance where $SD_{1}^2$ and $SD_{2}^2$ is the sample standard deviation and $N_{1}$ and $N_{2}$ is the sample size for group 1 and 2, respectively.

@@

## **Meta-Analytic Models to Control for Nuisance Heterogeneity and Investigate Heterogeneous Effects Within and Across Studies**

It will not always be possible to derive an effect size that completely controls for nuisance heterogeneity at the effect size level. This limitation is partly because treatment effects may not be simple linear or exponential functions of the mean response, making the effect sizes we discuss above unsuitable [e.g., thermal performance curves; @Noble2018]. For example, dosages applied to treatments can follow quite complex non-linear relationships in relation to some mean response (e.g., quadratic relationships). When this occurs, we recommend comparative physiologists apply meta-regression approaches, controlling for nuisance heterogeneity as a moderator. In many cases, this will be the easiest, or even preferred, approach because exploring drivers of heterogeneity is of prime interest in most meta-analyses anyway. Nonetheless, there are some simple statistical approaches a comparative physiologist can use to make the overall effects more easily interpretable, allowing one to control for nuisance heterogeneity, while also testing other biological moderators of interest [@Schielzeth2010]. Before diving into meta-regression models, we will first overview the meta-analytic models commonly used.

@@

A common goal of meta-analysis is to obtain an overall meta-analytic mean effect size estimate and some form of uncertainty around that mean, like a 95% confidence interval (CI). Including a prediction interval (PI) can also be useful (discussed below) [@Nakagawa2021c]. Estimating an overall mean and 95% CI is generally done with a simple random-effects or multilevel meta-analytic model that accounts for effect size sampling error [@Nakagawa2017; @NakagawaSantos2012]. More often than not, meta-analyses in comparative physiology will have complex hierarchical structure. For example, multiple effect sizes may be taken from a single study on traits measured on the same individuals or from many different species that share an evolutionary history [@Cinar2021; @Noble2017]. As such, it is far more likely that comparative physiologists will need to apply a multilevel meta-analytic model to control for the various sources of non-independence [@Cinar2021; @NakagawaSantos2012; @Song2021; @Nakagawa2021; @Noble2017]. Multilevel models are also extremely useful in describing the various drivers of effect size variance (a feature we describe below). For the purpose of our discussion, we will assume one needs to use more advanced multilevel models. Having said that, in some instances, it may be sufficient for a meta-analysis to make use of traditional random effects models. We refer readers to excellent reviews and books that overview the distinctions between common (fixed) effect, random-effect and multilevel models and when one might want (or not want) to apply these models [@Nakagawa2017; @NakagawaSantos2012; @Koricheva2013; @Borenstein2009; @Schmid2021].

@@

This model is a 'weighted' multilevel meta-analysis model because the estimation of the overall meta-analytic mean, $\beta_{0}$, is partially weighted by the inverse sampling variance of each effect size. Weighted meta-analytic models are important to use because they: 1) improve the precision on meta-analytic mean estimates by accounting for sampling variance; 2) allow for a formal analysis of heterogeneity; 3) and give precedence to higher quality studies (see below). From our survey, `r with(weighted, weighted[meta_analysis_weighted == "weighted model","percent"])*100`% of studies used a weighted meta-analytic model; however, of the studies self described as 'meta-analyses', `r with(weighted_meta, weighted_meta[meta_analysis_weighted == "unweighted model","percent"])*100`% (N = `r with(weighted_meta, weighted_meta[meta_analysis_weighted == "unweighted model","n"])` of `r with(weighted_meta, sum(weighted_meta[,"n"]))`) did not account for sampling variance. Many studies used 'Linear Regression' (including linear and generalised linear mixed effects models) as the main type of model fit to the data (Figure \@ref(fig:figsurv)C). Of the studies using weighted meta-analytic models `r with(het_reported, sum(het_reported[1:2,"percent"])*100)`% (N = `r with(het_reported, sum(het_reported[1:2,"n"]))` meta-analyses) did not report any measure of heterogeneity, which is unfortunate given how important these measures are to interpreting effects.

@@

#### *Multilevel Meta-regression to Understand Variation, Account for Nuisance Heterogeneity and Improve Effect Interpretation*

Heterogeneity among effects that result from nuisance heterogeneity can be dealt with using multilevel meta-regression models. Multilevel meta-regression models include all the same random effects that we have already discussed, but they also include fixed effects (i.e., predictors or moderators) that attempt to explain changes in the mean effect size. In other words, we can modify our multilevel meta-analytic model and turn it into a multilevel meta-regression model as follows:

@@

Note that this formula does not include $\sigma^2_{m}$ as sampling error variance is assumed to be known in meta-analysis, as explained above. When including continuous moderators, such as dosage, temperature, salinity, and even pH, $R^2_{marginal}$ provides a formal means to assess just how much variation in effects is the result of nuisance heterogeneity. Other biological moderators of interest could also be included as fixed effects and $R^2_{marginal}$ can be calculated for each independently, or all together. Importantly, moderators could be those that explain variation at the effect size level (i.e., dosage, temperature) or variation at the study or species level (e.g., reproductive mode, endothermy, etc.). $R^2_{marginal}$, or how we like to think of it in this context as $R^2_{nuisance}$, may be a useful measure to help readers understand just how much effect variation can result from the specific treatment application. If the heterogeneity is high as a result of, say, dosage differences, then it is clear that the choice of dosage will have a critical impact on the effect of interest. Comparative physiologists interested in implementing these calculations can do so using our packages `orchaRd` [@Nakagawa2021c] or `metaAidR` (<https://github.com/daniel1noble/metaAidR>).

@@

Comparative physiologists are interested in meta-analysing effects across a diversity of species, experimental designs, and environments. Importantly, sampling variances for many effect sizes commonly used by comparative physiologists (e.g., CTmax, $Q_{10}$ etc) can be formalized and powerful meta-analytic models can be easily applied to these effect measures (Figure \@ref(fig:fig0)). While an overall meta-analytic mean may be of interest, more often than not, understanding the drivers of effect size heterogeneity will be the primary interest. However, factors such as temperature and dosage differences across studies may generate obvious sources of heterogeneity that may not be of prime interest. Instead, comparative physiologists might simply want to focus on biological drivers of effect size variability and want an effect size or analytic approach that allows them to remove these ‘nuisances’. Surveying the comparative physiology literature shows that nuisance heterogenity is common, and is often not completely dealt with in a meta-analysis. Here, we provide a set of tools to deal with nuisance heterogeneity (Figure \@ref(fig:fig1)) by reinventing standard effect sizes and/or using mean centered nuisance variables in meta-regression models. Estimating complex meta-regression models with a multitude of fixed effects might result, however, in more challenging model interpretation. As such, a combined approach might be more desirable. Nuisance heterogeneity can be dealt with by using an appropriate effect size and associated sampling variance in combination with meta-regression models that contain moderators that test biological hypotheses of interest. Indeed, this may even be a necessity to simplify the model. The set of tools we describe provide clarification around the importance and limitations of heterogeneity. We also hope to provide new ways to help comparative physiologists communicate effect measures more richly to guide our understanding and decisions around pressing global challenges.
